Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-11-04 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Corrected up to p. 9

An early unpublished draft of Section \ref{sec:newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Siders
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bjordal and H\'ajek were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{controversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
released, respectively, in Isabellle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of G\"odel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of G\"odel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled its
verification with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a failed attempt to formalize Avicenna's
argument in PVS as well.

@@

The natural deduction calculus proposed and used here has standard rules for propositional connectives and quantifiers, as shown in Figures \ref{fig:PropositionalRules} and \ref{fig:QuantifierRules}. The extension to classical logic is achieved by adding a rule for double negation elimination, shown in Figure \ref{fig:Classical}, and $\beta$-reduction\footnote{Rules for $\eta$-expansion \citep{BenzmuellerBrownKohlhase} are not included here, because they are not needed in the ontological arguments.} is handled by the $\beta$ rule shown in Figure \ref{fig:ReductionRules}.
Finally, modal operators are handled by special new rules that insert or remove formulas from boxes, as shown in Figure \ref{fig:NDK}.
A \emph{derivation} is a directed acyclic graph whose nodes are formulas and whose edges correspond to applications of the inference rules. A \emph{proof} of a formula $F$ is a derivation without open assumptions and having $F$ as root not inside any box.

In every inference rule, (sub-)formulas in the conclusion correspond\footnote{This correspondence could be defined formally, but it would be tedious to do so.} to formulas in the premises (even in the case of quantifier rules, where the formulas differ due to substitution). For any expression occurring both in a formula in the premise and in its corresponding (sub-)formula in the conclusion, the occurrence in the premise is said to be an \emph{ancestor} of the occurrence in the conclusion, and the occurrence in the conclusion is said to be a \emph{descendant} of the occurrence in the premise.

@@

\begin{center}
\textbf{eigen-box condition:}\\ 
An modal inference is said to \emph{access}\\
 the box immediately above or below it. \\
$\nec_I$ and $\pos_E$ are \emph{strong} modal rules: \\
$\omega$ must be a fresh name for the box they access \\ 
(in analogy to the eigen-variable condition for strong quantifier rules). \\
Every box must be accessed by \emph{exactly one} strong modal inference. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are created.
\end{center}

@@

\begin{footnotesize}
\begin{prooftree}
\AXC{}\RightLabel{$1$}
\UIC{$P(\alpha)$}\RightLabel{$\imp_I^1$}
\UIC{$P(\alpha) \imp P(\alpha)$}
\AXC{}\RightLabel{$2$}
\UIC{$P(\alpha)$}\RightLabel{$\imp_I^2$}
\UIC{$P(\alpha) \imp P(\alpha)$}\RightLabel{$\wedge_I$}
\UIC{$(P(\alpha) \imp P(\alpha)) \wedge (P(\alpha) \imp P(\alpha))$} \dottedLine
\BIC{$P(\alpha) \biimp P(\alpha)$}\RightLabel{$\forall_I$}
\UIC{$\forall P. P(\alpha) \biimp P(\alpha)$}\dottedLine
\UIC{$\alpha=\alpha$} \RightLabel{$\forall_I$}
\UIC{$\forall x. x=x$} 
\end{prooftree}
\end{footnotesize}

>>>>

An early unpublished draft of Section \ref{sec:newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Siders
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bjordal and H\'ajek were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{controversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
released, respectively, in Isabelle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of G\"odel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of G\"odel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled its
verification with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a failed attempt to formalize Avicenna's
argument in PVS as well.

@@

The natural deduction calculus proposed and used here has standard rules for propositional connectives and quantifiers, as shown in Figures \ref{fig:PropositionalRules} and \ref{fig:QuantifierRules}. The extension to classical logic is achieved by adding a rule for double negation elimination, shown in Figure \ref{fig:Classical}, and $\beta$-reduction\footnote{Rules for $\eta$-expansion \citep{BenzmuellerBrownKohlhase} are not included here, because they are not needed in the ontological arguments studied here.} is handled by the $\beta$ rule shown in Figure \ref{fig:ReductionRules}.
Finally, modal operators are handled by special new rules that insert or remove formulas from boxes, as shown in Figure \ref{fig:NDK}.
A \emph{derivation} is a directed acyclic graph whose nodes are formulas and whose edges correspond to applications of the inference rules. A \emph{proof} of a formula $F$ is a derivation without open assumptions and having $F$ as root not inside any box.

In every inference rule, (sub-)formulas in the conclusion correspond\footnote{This correspondence could be defined formally, but it would be tedious to do so.} to formulas in the premises (even in the case of quantifier rules, where the formulas differ due to substitution). For any expression occurring both in a formula in the premise and in its corresponding (sub-) formula in the conclusion, the occurrence in the premise is said to be an \emph{ancestor} of the occurrence in the conclusion, and the occurrence in the conclusion is said to be a \emph{descendant} of the occurrence in the premise.

@@

\begin{center}
\textbf{eigen-box condition:}\\ 
a modal inference is said to \emph{access}\\
 the box immediately above or below it. \\
$\nec_I$ and $\pos_E$ are \emph{strong} modal rules: \\
$\omega$ must be a fresh name for the box they access \\ 
(in analogy to the eigen-variable condition for strong quantifier rules). \\
Every box must be accessed by \emph{exactly one} strong modal inference. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are created.
\end{center}

@@

\begin{footnotesize}
\begin{prooftree}
\AXC{}\RightLabel{$1$}
\UIC{$P(\alpha)$}\RightLabel{$\imp_I^1$}
\UIC{$P(\alpha) \imp P(\alpha)$}
\AXC{}\RightLabel{$2$}
\UIC{$P(\alpha)$}\RightLabel{$\imp_I^2$}
\UIC{$P(\alpha) \imp P(\alpha)$}\RightLabel{$\wedge_I$}
\BIC{$(P(\alpha) \imp P(\alpha)) \wedge (P(\alpha) \imp P(\alpha))$} \dottedLine
\UIC{$P(\alpha) \biimp P(\alpha)$}\RightLabel{$\forall_I$}
\UIC{$\forall P. P(\alpha) \biimp P(\alpha)$}\dottedLine
\UIC{$\alpha=\alpha$} \RightLabel{$\forall_I$}
\UIC{$\forall x. x=x$} 
\end{prooftree}
\end{footnotesize}
